Kindly Bent to Free Us

### 2019-07-10 kindly/lang.tex
.-.. regions

\subsubsection{Copying and Splitting}
\label{sdtyping:split}

Linear and affine types also control the copy of variables.
To demonstrate how this is achieve in \lang, we show the
{\sc Pair} typing rule in \cref{sdtyp:pair}.
In the {\sc Pair} rule, we split the type environment $\E$ in two parts,
$\E_1$ and $\E_2$, which are used to typecheck the components of the pair.
We note the splitting $\lsplit{C}{\E}{\E_1}{\E_2}$ where $C$ is a constraint
that must be respected for the split to be valid.
The main splitting rules are shown in \cref{sdtyp:split}. Each
rule applies to a binding for a given variable. For instance, the
{\sc Both} rule indicates that if we can prove that $x$ is unrestricted, we
can duplicate it in both environments.
Similarly, since immutable borrows are always unrestricted, the
{\sc Borrow} rule allows to duplicate them arbitrarily.
The {\sc Left} and {\sc Right} rule allow to move a binding
to the left or right environment.
The {\sc Susp} and {\sc SuspB} allow to split off suspended bindings to
the left environment while conserving access to the binding on the right.
Such suspended binding can later be turned
into a borrow inside a region. The asymmetry here must always follow
the order of execution of the language: in \affe, execution is from left to
right, which means that a resource can be used as a borrow on the left
and then as a full resource on the right, assuming that the borrow does not
escape. The {\sc Susp} rule work with a full resource while the {\sc SuspB}
rule allow to split off an immutable suspended binding based off a borrow.

Using these various rules, we can distribute the bindings every time an
expression has multiple sub-expressions such as pairs, function applications
and let bindings. For instance, in the
expression
$\letin{a}{\text{create}\ 10\ x}
{\introPair{\region[{}]{\Sone x\IBORROW}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
we use the rule {\sc Susp} to detach a borrow on the left of the pair.
Naturally, we must here ensure that the borrow on the left of the pair does not
escape, which brings us to the typing rules for regions.

\begin{figure}[!h]
  \centering
  \begin{subfigure}{0.35\linewidth}
    \begin{mathpar}
      \ruleSDPair
    \end{mathpar}
    \caption{The {\sc Pair} rule}
    \label{sdtyp:pair}
  \end{subfigure}\hfill
  \begin{subfigure}{0.6\linewidth}
    \centering
    \begin{tabular}
      {@{}>{$}r<{$}@{ $\vdash_e$ }
      >{$}c<{$}@{ $=$ }
      >{$}c<{$}@{ $\ltimes$ }
      >{$}c<{$}r}

@@

@@

      {\Cempty}&
      {\bvar{\borrow x} \schm}&{\svar[\IBORROW] x \schm^n}&{\bvar{\borrow x} \schm}
      &SuspB\\[2mm]

      {\Cempty}&
      {\bvar{\borrow[i]{x}}{\schm}}&
      {\bvar{\borrow[i]{x}}{\schm}}&{\bvar{\borrow[i]{x}}{\schm}}
      &Borrow\\

    \end{tabular}
    \caption{Splitting rules for bindings}
    \label{sdtyp:split}
  \end{subfigure}
\end{figure}

\subsubsection{Regions}
\label{sdtyping:regions}

In order to provide good support for imperative programming, \affe allows
to borrow resources.
Such borrows are constrained to a given syntactic region. We note
$\region{\Sone x\BORROW}{e}$ the region which allow borrows of $x$ to be used in $e$
and where $n$ is the nesting of the region.
Borrows can not be used outside of a corresponding region.
These remarks lead to an
algorithm for automatic region annotation: start from each borrow, create
a region, and extend it as much as possible while respecting other borrows
and binders.
A complete specification of the algorithm is provided in \cref{regionannot}.
In the rest of this section, we will only consider terms
with fully annotated regions.

The main idea of region typing (implemented by the {\sc Region} rule)
is to replace suspended bindings by borrow bindings, typecheck the body
of the region, and finally ensure that the borrow made available inside
the region does not leak outside.
This last check is done by using integer indices that correspond to the nesting
level of the region. The kind of borrow is annotated with the integer
corresponding to their region, and a final constraint ensure that
the return type of the region does not contain such borrow.

We illustrate this on the expression $\region{\Sone x\IBORROW}{f(\borrow[\IBORROW]{c})}$
when $c$ is a linear channel.
The first step is to check that
$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
%
This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
whose main rules are shown in \cref{selectrules:region}.
In our case, we consider an immutable binding, so we replace the suspended
binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
$\kun_n$).

\begin{figure}[!h]
  \begin{minipage}{0.4\linewidth}
    \centering
    \begin{mathpar}
      \ruleSDRegion
    \end{mathpar}
    \caption{The {\sc Region} rule}
    \label{selectrules:region}
  \end{minipage}\hfill
  \begin{minipage}{0.6\linewidth}
    \centering
    % \begin{tabular}
    %   {@{}>{$}r<{$}@{ $\vdash_e$ }
    %   >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
    %   >{$}l<{$}
    %   r}

@@

>>>>

\subsubsection{Copying and Splitting}
\label{sdtyping:split}
\begin{figure}[tp]
  \centering
  \begin{minipage}{0.35\linewidth}
    \begin{mathpar}
      \ruleSDPair
    \end{mathpar}
    \caption{The {\sc Pair} rule}
    \label{sdtyp:pair}
  \end{minipage}\hfill
  \begin{minipage}{0.6\linewidth}
    \centering
    \begin{tabular}
      {@{}>{$}r<{$}@{ $\vdash_e$ }
      >{$}c<{$}@{ $=$ }
      >{$}c<{$}@{ $\ltimes$ }
      >{$}c<{$}r}

@@

      {\Cempty}&
      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&
      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&{\bvar{\borrow[\IBORROW]{x}}{\schm}}
      &Borrow\\[2mm]

@@

      {\Cempty}&
      {\bvar{\borrow x} \schm}&{\svar[\IBORROW] x \schm^n}&{\bvar{\borrow x} \schm}
      &SuspB\\

    \end{tabular}
    \caption{Splitting rules for bindings}
    \label{sdtyp:split}
  \end{minipage}
\end{figure}

The {\sc Pair} typing rule in \cref{sdtyp:pair} demonstrates how \lang
deals with duplication and dropping of values.
The splitting $\lsplit{C}{\E}{\E_1}{\E_2}$ in the rule decomposes the
type environment $\E$ in two parts, $\E_1$ and $\E_2$, which are used
to typecheck the components of the pair. 
The constraint $C$ characterizes a valid  splitting.

\cref{sdtyp:split} shows the action of the splitting rules on single
bindings. If $x$'s type is unrestricted, then
the {\sc Both} rule indicates that we can duplicate it.
Similarly, unrestricted borrows can be duplicated according to rule
{\sc Borrow}.
The {\sc Left} and {\sc Right} rules are always applicable and move a binding
either to the left or right environment.
The rules {\sc Susp} and {\sc SuspB} split off suspended bindings to
the left while conserving access to the binding on the right.
A suspended binding can later be turned
into a borrow inside a region. Splitting of suspended bindings is
asymmetric. It must follow the order of execution from left to right,
which means that a resource can be used first as a borrow on the left 
and then later as a full resource on the right. The {\sc Susp} rule
works with a full resource while the {\sc SuspB} 
rule splits off an immutable suspended binding from a borrow.

Splitting applies whenever an
expression has multiple subexpressions:  pairs, function applications,
and let bindings. For instance, in the
expression
$\letin{a}{\text{create}\ 10\ x}
{\introPair{\region[{}]{\Sone x\IBORROW}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
the rule {\sc Susp} splits off a borrow from  the resource
\lstinline/a/ to use it in the left component of the pair.
As usual, a borrow cannot be active in the same scope as its resource.
The \emph{region} around its use ensures that the borrow in the left component does not
escape, which brings us to next topic.

\subsubsection{Regions}
\label{sdtyping:regions}
\begin{figure}[tp]
  \begin{minipage}{0.4\linewidth}
    \centering
    \begin{mathpar}
      \ruleSDRegion
    \end{mathpar}
    \caption{The {\sc Region} rule}
    \label{selectrules:region}
  \end{minipage}\hfill
  \begin{minipage}{0.6\linewidth}
    \centering
    % \begin{tabular}
    %   {@{}>{$}r<{$}@{ $\vdash_e$ }
    %   >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
    %   >{$}l<{$}
    %   r}

@@

Borrowing is crucial to support an imperative programming style.
To guarantee the validity of a borrow, its lifetime must be properly contained in its
ancestors lifetime. \lang ensures proper nesting of lifetimes by using
regions. The expression $\region{\Sone x\BORROW}{e}$ indicates a 
region at nesting level $n$ in which a $\BORROW$-borrow can be taken of $x$.

The typing for a region (see the {\sc Region} rule in \ref{selectrules:region})
replaces suspended bindings by borrow bindings, typechecks the body
of the region, and finally ensures that the borrow does not leak outside.
This last check is done with integer indices that correspond to the nesting
level of the region. The kind of a borrow is indexed with the level
corresponding to its region and the constraint $\Cleq{\tau}{\klin_{n-1}}$ ensures that
the return type of the region must live at some enclosing, lower level.

As an example, consider the expression $\region{\Sone x\IBORROW}{f(\borrow[\IBORROW]{c})}$
where $c$ is a linear channel.
The first step is to check that
$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
%
This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
whose main rules are shown in \cref{selectrules:region}.
In our case, we consider an immutable binding, so we replace the suspended
binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
$\kun_n$).

These remarks lead to an
algorithm for automatic region annotation: start from each borrow, create
a region, and extend it as much as possible while respecting other borrows
and binders.
A complete specification of the algorithm is provided in \cref{regionannot}.
In the rest of this section, we will only consider terms
with fully annotated regions.
